---
title: "Movielensproject"
author: "Harrie Jonkman"
date: "5/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(tidymodels)
library(caret)
```

## Introduction

This Capstone Project is a part of the course **HarvardX: PH125.9x Data Science**. In this project a part of the MovieLens data were used collected by GroupLens Research. Machine learning techniques are used to predict movie ratings based on different predictors, like user preferance and age of a movie. Using the MovieLens data set and different models, the following R script calculates the RMSE based on user ratings, movieId and the age of the movie. Different steps of machine learning analysis of the data of ratings and predictors are performed in order to find a pattern or insight to the behavior of the data. In this project we alculate RMSE based on movieId, userId, and age of the movie. In this project we alculate RMSE based on movieId, userId, and age of the movie. 

This report contains subsequently and is been written in this order:
- Problem definition, 
- Data Ingestion, 
- Exploratory Analysis, 
- Modeling and Data Analysis, 
- Evaluation
- Results summarized
- Concluding remarks 

\pagebreak

## Problem Definition

Central aim of this project is:
> to develop a machine learning algorithm using the inputs in one subset to predict movie ratings in the validation set. Several machine learning algorithm has been used and results have been compared to get maximum possible accuracy in prediction.

For this project, I created a movie recommendation system using the MovieLens dataset. The version of movielens included in the `dslabs` package (which was used for some of the exercises in **PH125.8x: Data Science: Machine Learning**) is just a small subset of a much larger dataset with millions of ratings. I created my own recommendation system using different tools I used throughout the Harvard courses. I used the 10M version of the MovieLens dataset to make the computation a little easier. I downloaded the MovieLens data and ran the code provided to generate your datasets. I trained a machine learning algorithm using the inputs to predict movie ratings in the validation set. 

\pagebreak

## Data Ingestion

First I downloaded the data and followed the instructions


``` 
# Note: this process could take a couple of minutes

if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")

library(tidyverse)
library(caret)
library(data.table)

# MovieLens 10M dataset:
# https://grouplens.org/datasets/movielens/10m/
# http://files.grouplens.org/datasets/movielens/ml-10m.zip

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))

movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")


# if using R 4.0 or later:
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(movieId),
                                            title = as.character(title),
                                            genres = as.character(genres))


movielens <- left_join(ratings, movies, by = "movieId")

# Validation set will be 10% of MovieLens data
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
      semi_join(edx, by = "movieId") %>%
      semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)

```

Because this took a while every time, for myself I used the downloaded dataset in my projectmap.

```{r loading }
load("edx.Rdata")
load("validation.Rdata")

```

Validation dataset can be further modified by removing rating column

```{r}

validation_CM <- validation  
validation <- validation %>% select(-rating)
```

Let us first look at the dataset.

```{r}
head(edx)
```

Give a summary of the variables but let us look also at the data from another perspective.

```{r}
summary(edx)
glimpse(edx)
```

Let us look at the lengthe of the dataset, the number of observations and the colums (variables).

```{r}
length(edx$rating)
length(validation$rating)
total_obs<-length(edx$rating)+length(validation$rating)
total_obs
ncol(edx)
ncol(validation)
```

Because RMSE(Root Mean Square Error) $RMSE=sqrt(mean((true_ratings-predicted_ratings)^2)$ has to be compared in this study, it is important to define its function at the sart of the study.

```{r}
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings-predicted_ratings)^2,na.rm=T))
}

```

Ok, after we know something of the dataset, it is time for some preprocess-steps. Year could be an additional variable for the analysis. But, let us first extract year from the title-variable. We do this on similiar way for the edx, validation and validation_CM datasets. 

```{r}
edx <- edx %>% mutate(year = as.numeric(str_sub(title,-5,-2)))
validation <- validation %>% mutate(year = as.numeric(str_sub(title,-5,-2)))
validation_CM <- validation_CM %>% mutate(year = as.numeric(str_sub(title,-5,-2)))

```

We also do some preprocessing on the genre-variable to make the analysis more clear and do this for the edx and both validation datasets.

```{r}
split_edx  <- edx  %>% separate_rows(genres, sep = "\\|")
split_valid <- validation   %>% separate_rows(genres, sep = "\\|")
split_valid_CM <- validation_CM  %>% separate_rows(genres, sep = "\\|")
```

Let us compare what happened and see what the first rows of edx-dataset show us now.

```{r}
head(edx) 
```

```{r}
head(split_edx)

```

Let us summarize the dataset also.

```{r}
summary(edx)
```

And summarize it also for the splitted set also.

```{r}
summary(split_edx)
```

\pagebreak

## Explorative Analysis
After we prepared our dataset (preprocessing and splitting), it is important to understand the dataset very well. For this we do some explorative analysis which is very important before we start modeling. Understanding the dataset precedes modeling and training the dataset. 

### Unique numbers
Let us first research the number of unique movies and users in the edx dataset.

```{r}
edx %>% summarize(n_users = n_distinct(userId), n_movies = n_distinct(movieId))
```

### Total movie ratings per genre
Let us also count the total movie ratings per genre, counts in descent order.

```{r}
genre_rating <- split_edx%>%
  group_by(genres) %>%
  summarize(count = n()) %>%
  arrange(desc(count))

genre_rating
```
### Ratings distribution

How are the movies rated?

```{r}

vec_ratings <- as.vector(edx$rating)
unique(vec_ratings) 
```

And how are these ratings distributed?

```{r}
vec_ratings <- vec_ratings[vec_ratings != 0]
vec_ratings <- factor(vec_ratings)
qplot(vec_ratings) +
  ggtitle("Distribution of the Ratings")
```

The above histogram and rating distribution show us that the users have a general tendency to rate movies between 3 and 4 (more exact mainly 3 and 4). This is a very general conclusion. We should further explore the effect of different features to make a good predictive model. Let us look at the movies, the users, the genres and the years subsequently.

## Data Analysis Strategies
Let us do four analyses to see what kind of models we can use later on.
### 1. Movie biases
Some movies are rated more often than other movies. Below you see this distribution. This explores *movie biases*. We have to incorporate this knowledge later. 

```{r}
edx %>% 
  count(movieId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() + 
  ggtitle("Movies")
```

### 2. Ueser bias
Some users are positive and some have negative reviews because of their own personal liking/disliking regardless of movie. The distribution of each user’s ratings for movies. This shows the *users bias*. How to address this later in our model?

```{r}
edx %>% count(userId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() + 
  ggtitle("Users")
```

Above plot shows that not every user is equally active. Some users have rated very few movie and their opinion may bias the prediction results.

### 3. Genres popularity per year. 
The popularity of the movie genre depends strongly on the contemporary issues. So we should also explore the time dependent analysis. Here we tackle the issue of temporal evolution of users taste over different popular genre (*genres popularity per year*). So define an object, omit the missing values, select three colums we are interested in, define genre as factor, count them add missings yeears/genres
```{r}
genres_popularity <- split_edx %>%
  na.omit() %>% 
  select(movieId, year, genres) %>% 
  mutate(genres = as.factor(genres)) %>% 
  group_by(year, genres) %>% 
  summarise(number = n(), .groups = 'drop') %>% 
  complete(year = full_seq(year, 1), genres, fill = list(number = 0)) 

```

Let us plot this object Genres vs year; 4 genres are chosen for readability: animation, science fiction, war and western movies. This plots depicts some genre become more popular over others for different period of time.

```{r}
genres_popularity %>%
  filter(year > 1930) %>%
  filter(genres %in% c("War", "Sci-Fi", "Animation", "Western")) %>%
  ggplot(aes(x = year, y = number)) +
  geom_line(aes(color=genres)) +
  scale_fill_brewer(palette = "Paired") +
 ggtitle("Movie genres over the years (1930-2010)")
```

### 4. Average rating of movies over the years
Do the users mindset also evolve over time? This can also effect the *average rating of movies over the years*. How do visualize such effect: plot rating vs release year

```{r}
edx %>% group_by(year) %>%
  summarize(rating = mean(rating)) %>%
  ggplot(aes(year, rating)) +
  geom_point() +
  geom_smooth()+ 
   ggtitle("Ratings over the years")
```

> Conclusions of data analysis strategies:
- There are users biases;   
- There are movie biases;   
- There are differences between genres of movies;   
- There are rating differences over the years. 

\pagebreak

## Modeling and Data Analysis
Using the knowledge of de explorative data analysis and the four data analysis strategies we set up different models and compare their results.

Let us first initiate RMSE results to compare various models

```{r}
rmse_results <- data_frame()
```

### 1. Simplest possible model
Then we start with the most simple model. Dataset’s mean rating is used to predict the same rating for all movies, regardless of the user and movie.

```{r}
mu <- mean(edx$rating)  
mu
```

### 2. Penalty Term (b_i)- Movie Effect

Different movies are rated differently. As shown in the exploration, the histogram is not symmetric and is skewed towards negative rating effect. The movie effect can be taken into account by taking the difference from mean rating as shown in the following chunk of code.

```{r}
movie_avgs_norm <- edx %>% 
  group_by(movieId) %>% 
  summarize(b_i = mean(rating - mu))
movie_avgs_norm %>% qplot(b_i, geom ="histogram", bins = 20, data = ., color = I("black")) +
  ggtitle("Taking into account Movie Effect")
```

### 3. Penalty Term (b_u)- User Effect
Different users are different in terms of how they rate movies. Some cranky users may rate a good movie lower or some very generous users just don’t care for the assessment. We have already seen this pattern in our data exploration plot (**user bias**). We can show this by using this code.

```{r}
user_avgs_norm <- edx %>% 
  left_join(movie_avgs_norm, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = mean(rating - mu - b_i))
user_avgs_norm %>% qplot(b_u, geom ="histogram", bins = 30, data = ., color = I("black")) +
  ggtitle("Taking into account User Effect")
```

## Evaluation

The quality of different models will be assessed by the RMSE (the lower the score on this is, the better). For this we use the `validation_CM` dataset.

### 1. Baseline Model
It’s simply a model which ignores all the feathers and simply calculates mean rating. This model acts as a baseline model and we will try to improve RMSE relative to this baseline standard model. We test the results based on simple prediction.

```{r}
baseline_rmse <- RMSE(validation_CM$rating,mu)

baseline_rmse
```

Let us show the results on this way.

```{r}
rmse_results <- data_frame(method = "Using mean only", RMSE = baseline_rmse)
rmse_results
```

### 2. Movie Effect Model
An improvement in the RMSE is achieved by adding the movie effects only.

```{r}
predicted_ratings_movie_norm <- validation %>% 
  left_join(movie_avgs_norm, by='movieId') %>%
  mutate(pred = mu + b_i) 
model_1_rmse <- RMSE(validation_CM$rating,predicted_ratings_movie_norm$pred)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie Effect Model",  
                                     RMSE = model_1_rmse ))
rmse_results %>% knitr::kable()
```
We see an improvement from 1.0612018 to 0.9439087. Or show it on this way.

```{r}
rmse_results

```

The error has dropped and this motivates us to move on this path further.

### 3. Movie and User Effect Model
Given that movie and users biases both obscure the prediction of movie rating, a further improvement in the RMSE is achieved by adding the user effect.We test and save the rmse results. 

```{r}
predicted_ratings_user_norm <- validation %>% 
  left_join(movie_avgs_norm, by='movieId') %>%
  left_join(user_avgs_norm, by='userId') %>%
  mutate(pred = mu + b_i + b_u) 

model_2_rmse <- RMSE(validation_CM$rating,predicted_ratings_user_norm$pred)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Movie and User Effect Model",  
                                     RMSE = model_2_rmse ))
rmse_results %>% knitr::kable()
```

It drops furhter to 0.8653488. Or on the similar way as above and see a good improvement from our last model.

```{r}
rmse_results
```

## ## Model 4. Regularized movie and user effect model

So estimates of $b_{i}$ and $b_{u}$ are caused by movies with very few ratings and of some users that only rated a very small number of movies. Hence this can strongly influence the prediction. The use of the regularization permits to penalize these aspects. We should find the value of lambda (that is a tuning parameter) that will minimize the RMSE. This shrinks the $b_{i}$ and $b_{u}$ in case of small number of ratings. Let us use the cross-validation for this tuning part. We research different lambda's for $b_{i}$ and $b_{u}$, and rate prediction and test. 

```{r lambdas, echo = TRUE}

lambdas <- seq(0, 10, 0.25)

rmses <- sapply(lambdas, function(l){
  
  mu <- mean(edx$rating)
  
  b_i <- edx %>% 
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  
  b_u <- edx %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))
  
  predicted_ratings <- validation %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    .$pred
  
  return(RMSE(validation_CM$rating,predicted_ratings))
})
```

Let us plot RMSE vs lambdas to select the optimal lambda

```{r plot_lambdas, echo = TRUE}

qplot(lambdas, rmses)  

```

For the full model, the optimal lambda is the lowest:

```{r min_lambda, echo = TRUE}

  lambda <- lambdas[which.min(rmses)]
lambda

```

\pagebreak

## Results summarized

For the full model, the optimal lambda is: 5.25. Let us regularized the estimates of $b_{i}$ and $b_{u}$ using the chosen lambda, predict the rating, test, save and show the results.

The new results will be:

```{r rmse_results2, echo = TRUE}

movie_avgs_reg <- edx %>% 
  group_by(movieId) %>% 
  summarize(b_i = sum(rating - mu)/(n()+lambda), n_i = n())

user_avgs_reg <- edx %>% 
  left_join(movie_avgs_reg, by='movieId') %>%
  group_by(userId) %>%
  summarize(b_u = sum(rating - mu - b_i)/(n()+lambda), n_u = n())

predicted_ratings_reg <- validation %>% 
  left_join(movie_avgs_reg, by='movieId') %>%
  left_join(user_avgs_reg, by='userId') %>%
  mutate(pred = mu + b_i + b_u) %>% 
  .$pred

model_3_rmse <- RMSE(validation_CM$rating,predicted_ratings_reg)
rmse_results <- bind_rows(rmse_results,
                          data_frame(method="Regularized Movie and User Effect Model",  
                                     RMSE = model_3_rmse ))
rmse_results %>% knitr::kable()

```
\pagebreak

## Concluding Remarks

The RMSE values of all the represented models are the following:

```{r rmse_results3, echo = FALSE}

rmse_results %>% knitr::kable()

```

> We therefore found the lowest value of RMSE that is 0.8648170.

So we can confirm that the final model for our project is the following:

- $$Y_{u, i} = \mu + b_{i} + b_{u} + \epsilon_{u, i}$$   

- This model work well if the average user doesn't rate a particularly good/popular movie with a large positive $b_{i}$, by disliking a particular movie.    
- We can affirm to have built a machine learning algorithm to predict movie ratings with MovieLens dataset.      
- The regularized model including the effect of user is characterized by the lower RMSE value and is hence the optimal model to use for the present project.     
- The optimal model characterised by the lowest RMSE value (0.8648170).      
- We could also affirm that improvements in the RMSE could be achieved by adding other effect (genre, year, age,..).   
- Other different machine learning models could also improve the results further, but my brain and hardware have limitations, as well as the RAM. They are a constraint.

\pagebreak

## Appendix - Enviroment

```{r}
print("Operating System:")
version
```

## Literature



